```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```

## Tips at Restaurants

In this homework we will use the `tips` data set. This data set is part of the `reshape2` package. You can load the data set by executing the command:
```{r}
library(reshape2)
data(tips, package="reshape2")
```
If you do not have available the package `reshape2`, issue `install.packages('reshape2')` to install it. The information contained in the data is collected by one waiter, who recorded over the course of a season information about each tip he received working in one restaurant. See `?tips` for a description of all of the variables. 

**Submission instructions**: Create a folder named `ds202_hw3`, and name the RMarkDown file `hw3.Rmd`  which should include your solutions under this folder. For submission, create a GitHub repository named `ds202_hw3` under your GitHub account, and push both `hw3.Rmd` and the knitted `hw3.html` before the deadline. I will assume you use the same GitHub username as for your HW2 submission. The latest version of your homework appearing on GitHub before the deadline is graded. *It is crucial to follow the exact folder structure and file names*, so that your homework can be reproduced and graded by an automated script. This homework is **due on Feb 19** before class.

### Lets take a look at the layout of the data

```{r}
head(tips)
```

1.  How many parties did the waiter serve? Store the number in `numParty` and print. 

```{r}
numParty = nrow(tips)
print(numParty)
```

2. What are the types of variables in the data frame `tips`? Include the code and also explain verbally.

```{r}
print(str(tips))
```

`toal_bill` and `tip` are `num` types, or floating point, `sex`, `smoker`, `day`, `time` are all `factor` types which mean that they're basically enums or `int` with labels attached to the values. Finally, `size` is an `int` which is self explanatory.

3. Create a vector named `day` containing the day of visits in `tips`. The factor levels should be ordered from Thursday to Sunday. Print the variable.

```{r}
day = factor(tips$day, levels=c('Thur', 'Fri', 'Sat', 'Sun'))
print(day)
```

4. Create a data frame named `female5` containing the meal paid by a female payer in a party with size greater than or euqal to 5. Print the data frame.

```{r}
female5 = tips[tips$sex == 'Female' & tips$size >= 5, ]
print(female5)
```

5. How does the tipping amount (`tip`) depend on the overall bill (`total_bill`)? Use the `ggplot2` package to make a chart. Describe the relationship in words.  

```{r}
ggplot(tips, aes(x=total_bill, y=tip, color=tip)) + geom_point()
```

There is a semi-linear relationship between `total_bill` and `tip` with most values sitting around the 20% mark, and many going below, and few going above.

6. Describe at least two types of anomalies in the previous plot. What do they mean?

Two types of anomalies would be the outliers, which are most likely from either poor/outstanding performance on the waiter's part, or maybe just a generous customer. And there are a couple of tip points that generate horizontal lines meaning that the denomination is really common as a tip among several bill totals; this is interesting because it could stem from several reasons:

- Does the bill total make exactly that amount of change when rounded up to a $5 bill?
- Is that an even dollar denomination?
- Is it somebody's favourite number?

It could be any of these things, it's hard to tell for certain.

7. Introduce a variable `tiprate` into the data set which stands for the rate of tips. What is the average rate for tips? Explain verbally. 

```{r}
tips$tiprate = tips$tip / tips$total_bill
mean(tips$tiprate)
```

Not sure how to explain this verbally, the average tip rate is 16%?

8. Make a boxplot of the tip rate by time. The x-axis should be ordered by lunch and then dinner. Use `ggplot2` to make a chart. Verbally explain the chart.

```{r}
ggplot(tips, aes(x=factor(time, levels=c('Lunch', 'Dinner')), y=tiprate)) + geom_boxplot() + xlab('Time') + ylab('Tip rate')
```

This chart shows that on average the tip rate does not depend on the time of day, but the distribution does. The standard deviation of tips is higher during Dinner than during Lunch.

9. How does smoking behavior and gender of the person who pays impact the relationship between tip and total bill? Find a visualization that incorporates all four variables. Interpret the result.

```{r}
ggplot(tips, aes(x=sex, y=tiprate, color=smoker)) + geom_point()
```

It would seem that smoking does seem to encompass all of the outliars indicating that smokers are mor likely to tip higher percentages. Women smokers are more extreme with their tips, either tipping really low, or really higher, whilst non-smoking women tend to stay right around the average. Men are just kind of all over the place.

10. Use ggplot2 to find a graphical summary of the relationship between day of the week and gender of the person paying the bill. What can you say about this relationship?

```{r}
ggplot(tips, aes(x=day, weight=tip, fill=sex)) + geom_bar()
```

That men tend to pay the bill in a disproportionate amount to women, except on Fridays, and slightly less disproportionate on Thursdays. People tip very little on Fridays.

Note: your submission is supposed to be fully reproducible, i.e. the TA and I will 'knit' your submission in RStudio. 
